Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-26 paper.tex
Pass through mutations section

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its usage of the ``succinct tree sequence''
 data stucture to represent simulation results.
The succinct tree sequence (usually abbreviated to tree sequence)
is a data structure for concisely encoding
genetic ancestry and sequence variation.
The basic concepts were introduced by \cite{kelleher2016efficient}
and originally developed as part of \msprime. We subsequently
extracted the tree sequence code that was independent of simulations
from \msprime\ into \tskit,
a standalone Python and C library~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation here.

% Encoding ancestry
Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges} (Fig.~\ref{fig-ts-example}).
A node in tree sequence terminology corresponds
to a single chromosome strand, and is associated with its birth time,
and other optional metadata. A node may optionally be
associated with an \emph{individual}, so that, for example, a diploid
individual will be associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within
some individual, and edges define genetic inheritence
relationships between these nodes. An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
generalises the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.

@@

% OK, back to tree sequences: what about sequence variation?
The final output of most population simulations is some representation
of the sequence variation among the specified samples. For coalescent
simulations, we usually have three steps (1) simulate
the genetic ancestry, and optionally output the resulting marginal trees;
(2) simulate sequence evolution conditioned on this ancestry by generating
mutations (see the \nameref{sec-mutations} section); and (3) output the
resulting sequences by percolating the effects of the mutations through
the trees. Information about the mutations
themselves---e.g., where they have occurred on the
trees---is usually not retained or made available for subsequent analysis.
In \msprime, however, we skip step (3), instead using \tskit's
data model of ancestry and mutations
to represent the simulated sequences.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
fully integrated part of \tskit's tree sequence data model, and genetic
variation is encoded by recording sites at which mutations
have occurred, and where each mutation at those sites
has occurred on the marginal tree.
Crucially, the sequences themselves are never stored, or indeed directly
represented in memory (unless explicitly decoded using \tskit\ APIs).
It may at first seem inconvenient to have only this indirect representation
of the genome sequences, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we would require $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space,
assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
For large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. Besides the huge
difference in storage efficiency, it is also often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data using \tskit\
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

@@

% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees
(see Fig.~\ref{fig-mutated-trees}).
A number of programs exist to place mutations on trees: for instance,
the classical \SeqGen\ program~\citep{rambaut1997seq}
supports a range of different models of sequence evolution,
and there are various extensions to the basic
models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
Partly for efficiency and partly in the interest of
simplicity for users (i.e., to avoid intermediate text format conversions),
population genetic simulators have tended to
include their own implementations of mutation simulation, with
most supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.

\begin{figure}
    \includegraphics{illustrations/unmutated_tree}
    \includegraphics{illustrations/mutated_tree}
    \caption{
        \label{fig-mutated-trees}
        Visualization of the separation between ancestry and mutation simulation:
        \texttt{sim\_ancestry} produces ancestry
        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
        relating 200 samples),
        and \texttt{sim\_mutations} adds mutations
        (\textbf{right;} fourteen nucleotide mutations have been added).
        Graphics produced with \tskit's \texttt{draw\_svg} method.
        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
    }
\end{figure}

% on the other hand, tskit is awesome.
Clean separation of the ancestry and mutation generation steps
is made possible by the \tskit\ library, whose mature and
efficient C and Python interfaces makes it straightforward to add
further information to an existing file, and to output the results.
As such, it is essentially a historical accident
that \msprime\ (whose main focus is on ancestry simulation)
contains code to simulate mutations: the mutation process could
be performed by a separate package, and there would be no performance
penalty for doing so.
Thanks to this interoperability, \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real data
using~\texttt{tsinfer}/\texttt{tsdate}~\citep{kelleher2019inferring,wohns2021unified}
or \texttt{Relate}~\citep{speidel2019method}. It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

For version 1.0 we have greatly extended the sophistication of
\msprime's mutation generation engine, achieving near feature-parity
with \SeqGen.
We  support a large number of mutation models, including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
A general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
can be easily and efficiently defined in user code.
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
The new \texttt{sim\_mutations} function
(see the \nameref{sec-sim-interface} section) defaults to a finite-sites
mutation model, but full support for the
original infinite sites model is also maintained.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

Generating simulations in \msprime\ is efficient.
Fig.~\ref{fig-mutations-perf} shows the time required to generate
mutations (using the default JC69 model) on
simulated tree sequences for a variety of mutation
rates as we vary the number of samples
(Fig.~\ref{fig-mutations-perf}A) and the sequence
length (Fig.~\ref{fig-mutations-perf}B).
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with 508,125 edges.
This efficiency for large numbers of trees is possible because
the tree sequence encoding allows us to generate mutations
on an edge-by-edge basis (see Fig.~\ref{fig-ts-example}), rather than
tree-by-tree and branch-by-branch as would otherwise be required.
In the above example from Fig.~\ref{fig-mutations-perf}B,
we would have to consider 273,887,838 branches
since there are 137,081 trees and 1,998 branches in each
tree ($\sim500$ times more work).
% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
Even if we have a tree sequence consisting of a single tree
(so that there is no advantage to working edge-by-edge),
\msprime's mutation generator is quite efficient.
For example, we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim$260,000 mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods because at this scale it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms.
It seems quite unlikely, however,
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.

There are many ways in which the mutation generation code
in \msprime\ could be extended. For example, we hope to add support for
microsatellites~\citep{mailund2005coasim},
codon models~\citep{arenas2007recodon}
and indels~\citep{cartwright2005dna,fletcher2009indelible}.

@@

Recombination is implemented in \msprime\ using Hudson's algorithm, which
works backwards in time tracking the
effects of common ancestor and recombination
events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
Common ancestor events combine the ancestral material of two lineages, and may
result in coalescences in the marginal trees. Recombination events
split the ancestral material for some lineage, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
us to simulate recombination hotspots like \texttt{msHOT}~\citep{hellenthal2007mshot}
and empirical recombination maps similar to other
simulators~\citep[e.g.][]{shlyakhter2014cosi2}.
The implementation of recombination in \msprime\ is extensively validated
against analytical results~\citep{hudson1983properties,kaplan1985use}
and results from \ms.

>>>>

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its usage of the ``succinct tree sequence''
 data stucture to represent simulation results.
The succinct tree sequence (usually abbreviated to ``tree sequence'')
is a data structure for concisely encoding
genetic ancestry and sequence variation.
The basic concepts were introduced by \cite{kelleher2016efficient}
and originally developed as part of \msprime. We subsequently
extracted the tree sequence code that was independent of simulations
from \msprime\ into \tskit,
a standalone Python and C library~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation here.

% Encoding ancestry
Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges} (Fig.~\ref{fig-ts-example}).
A node in tree sequence terminology corresponds
to a single chromosome strand, associated with its birth time
and other optional metadata. A node may be
associated with an \emph{individual}, so that, for example, a diploid
individual is associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within
some individual, and edges define genetic inheritence
relationships between these nodes. An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
generalises the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.

@@

% OK, back to tree sequences: what about sequence variation?
The final output of most population genetic simulations is some representation
of sequence variation among the specified samples. For coalescent
simulations, we usually have three steps: (1) simulate
the genetic ancestry, and optionally output the resulting marginal trees;
(2) simulate sequence evolution conditioned on this ancestry by generating
mutations (see the \nameref{sec-mutations} section); and (3) output the
resulting sequences by percolating the effects of the mutations through
the trees. Information about the mutations
themselves---e.g., where they have occurred on the
trees---is usually not retained or made available for subsequent analysis.
In \msprime, however, we skip step (3), instead using \tskit's
data model of ancestry and mutations
to represent the simulated sequences.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
fully integrated part of \tskit's tree sequence data model, and genetic
variation is encoded by recording sites at which mutations
have occurred, and where each mutation at those sites
has occurred on the marginal tree.
Crucially, the sequences themselves are never stored, or indeed directly
represented in memory (unless explicitly decoded using \tskit\ APIs).
It may at first seem inconvenient to have only this indirect representation
of the genome sequences, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we would require $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space,
assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
For large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. Besides the huge
difference in storage efficiency, it is also often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data using \tskit\
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

@@

% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees
(see Fig.~\ref{fig-mutated-trees}).
A number of programs exist to place mutations on trees: for instance,
the classical \SeqGen\ program~\citep{rambaut1997seq}
supports a range of different models of sequence evolution,
and various extensions to the basic
models have been proposed~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
Partly for efficiency and partly in the interest of
simplicity for users (i.e., to avoid intermediate text format conversions),
population genetic simulators have tended to
include their own implementations of mutation simulation, with
most supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a wide range of different models of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}. Thus, despite the logical separation between
the tasks of simulating ancestry and neutral sequence evolution,
the two have been conflated in practice.

\begin{figure}
    \includegraphics{illustrations/unmutated_tree}
    \includegraphics{illustrations/mutated_tree}
    \caption{
        \label{fig-mutated-trees}
        Visualization of the separation between ancestry and mutation simulation:
        \texttt{sim\_ancestry} produces ancestry
        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
        relating 200 samples),
        and \texttt{sim\_mutations} adds mutations
        (\textbf{right;} fourteen nucleotide mutations have been added).
        Graphics produced with \tskit's \texttt{draw\_svg} method.
        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
        \jkcomment{Why 200 samples? Seems like far too many to me, how about 5?}
    }
\end{figure}

Part of the reason for this poor record of software reuse and modularity is the
lack of standardised file formats, and in particular, the lack of common
library infrastructure to abstract the details of interchanging simulation
data. On the surface, it would appear that \msprime\ is also guilty of this
lack of modularity since it supports simulating both ancestry and mutations.
However, this is essentially a historical accident and because of the \tskit\
library, the mutation process in \msprime\ could just as easily
be performed in a separate package.
The mature and efficient C and Python interfaces for
\tskit\ makes it straightforward to add further information to
an existing file, and because of its efficient data interchange mechanisms,
there is no performance penalty for performing additional operations
in a different Python package.
Thanks to this interoperability, \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real data
using~\texttt{tsinfer}/\texttt{tsdate}~\citep{kelleher2019inferring,wohns2021unified}
or \texttt{Relate}~\citep{speidel2019method}. It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

As well as providing a new API (see the \nameref{sec-sim-interface} section)
which emphasises this logical split between ancestry and mutation simulation,
we have greatly extended the sophistication of
\msprime's mutation generation engine for version 1.0,
achieving near feature-parity with \SeqGen.
We  support a large number of mutation models, including the
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
A general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
can be easily and efficiently defined in user code.
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
The new \texttt{sim\_mutations} function
(see the \nameref{sec-sim-interface} section) defaults to a finite-sites
mutation model, but full support for the
original infinite sites model is also maintained.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

Simulating mutations in \msprime\ is efficient.
Fig.~\ref{fig-mutations-perf} shows the time required to generate
mutations (using the default JC69 model) on
simulated tree sequences for a variety of mutation
rates as we vary the number of samples
(Fig.~\ref{fig-mutations-perf}A) and the sequence
length (Fig.~\ref{fig-mutations-perf}B).
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with 508,125 edges.
This efficiency for large numbers of trees is possible because
the tree sequence encoding allows us to generate mutations
on an edge-by-edge basis (see Fig.~\ref{fig-ts-example}), rather than
tree-by-tree and branch-by-branch as would otherwise be required.
In the above example from Fig.~\ref{fig-mutations-perf}B,
we would have to consider 273,887,838 branches
since there are 137,081 trees and 1,998 branches in each
tree ($\sim500$ times more work).
% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
Even if we have a tree sequence consisting of a single tree
(so that there is no advantage to working edge-by-edge),
\msprime's mutation generator is still very efficient.
For example, we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim$260,000 mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods because at this scale it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms.
It seems quite unlikely, however,
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.

There are many ways in which the mutation generation code
in \msprime\ could be extended. For example, we hope to add support for
microsatellites~\citep{mailund2005coasim},
codon models~\citep{arenas2007recodon}
and indels~\citep{cartwright2005dna,fletcher2009indelible},
although some may also require upstream additions to \tskit's data model
which is currently based on the assumption of independent sites.

@@

Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
works backwards in time tracking the
effects of common ancestor and recombination
events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
Common ancestor events combine the ancestral material of two lineages, and may
result in coalescences in the marginal trees. Recombination events
split the ancestral material for some lineage, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
us to simulate recombination hotspots like \texttt{msHOT}~\citep{hellenthal2007mshot}
and empirical recombination maps similar to other
simulators~\citep[e.g.][]{shlyakhter2014cosi2}.
The implementation of recombination in \msprime\ is extensively validated
against analytical results~\citep{hudson1983properties,kaplan1985use}
and results from \ms.
